\chapter{Acknowledgements}
This Ph.D. journey would not be possible without these people to be acknowledged here.

First and foremost, I would like to express my deep gratitude towards my supervisor Prof. Lars K. Rasmussen. At the beginning of this journal, Lars remarked to me, 'Let us do something interesting'. This code revealed more power later as time went than it sounded at the beginning. In the long and challenging journey, it was the propulsion from interests that drove me all the way here. At the time that I lay these words to my thesis, I still feel the joy of doing what I do in research. Lars has been always supportive in advising me and teaching me the way to do research, though there were inevitably detours in research journeys. Whenever I was in moments of confusion and frustrations, Lars's positive attitude and humor brought me back to balance.
I would like to thank my supervisor Assoc. Prof. Ragnar Thobaben for his consistent support to me in my education. I benefited from every meeting with him. His comments on my thesis were always insightful input and to the point. The work with Ragnar was efficient and joyful. I was lucky to have Lars and Ragnar as my advisors on this journey.

I am grateful to Assoc. Prof. Saikat Chatterjee and Dr. Minh Th\`{a}nh Vu. There are always interesting ideas from Saikat's active mind. Fortunately, our offices are just several-step away, which allows quick discussions before the inspiration fades out. Thank Saikat for his sacrifice of weekends and holidays to help me improving my drafts.
I own the thanks to Minh for conveying many prohibitive problems in intuitive ways in our discussions, which gradually helps to form my own way of comprehending complex issues. I am impressed by how knowledgeable this mind is and how fast the mind processes problems. I learned a lot from him. I would like to express my sincere gratitude to Minh for the consistent guidance of my research.

I am indebted to Prof. Viktoria Fodor and Assoc. Prof. Chao Wang who shared valuable research experiences, were helpful, and supportive in collaborations. I am fortunate to have had joint work with Antoine Honor{\'e}, Baptiste Cavarec, Dr. Jing Yue, Dr. Jinliang Huang, Dr. Zuxing Li, and Dr. Nima N. Moghadam. I have benefited tremendously from interactions with them.

I would like to thank both Dr. Pol Del Aguila Pla and Xuechun Xu for their quick responses to my frequent requests for packages and dependencies on the computational servers. Thank Sina Molavipour for being my teaching partner. I would also like to thank Henrik Forssel for the help for the Swedish abstract.


I am fortunate to have fantastic colleagues at ISE. I really enjoyed the badminton activities with Alireza Mahdavi Javid, Ramana Reddy Avula, Xuechun, Zuxing, and Saikat. The relaxed time of lunches and fikas with Shaocheng Huang, Dr. Yu Ye, Boules Atef Mouris, Sahar Imtiaz, Dr. Guang Yang, Dr. Le Phuong Cao, Dr. Marie Maros, Dr. Germ\'{a}n Bassi, and Dr. Arun Venkitaraman were pleasant in office kitchens.

I am always grateful to my master study adviser Prof. Erwu Liu in Tongji for encouraging me to do Ph.D. research. This journey is challenging and fruitful.

I cannot thank enough to my mother, Zirong Li, who has never stopped encouraging me to pursue what I want, and has always been there for me, being supportive. \newline
\newline


\hfill
Dong Liu

\hfill
Stockholm, August 2020





%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:
